Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-05-31 2020/RSD_GPU/RSD2020.tex

\section{Multidimensional parallel global search algorithm}\label{Sec_GSA}
Consider the problem of finding the global minimum of the \(N\)-dimensional function \(\varphi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\).  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\varphi(y^*)=\min\{\varphi(y):y\in D\},
\end{equation}
\begin{equation}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty.
\end{equation}

@@

At the first iteration of the method, the trial is carried out at an arbitrary internal point \(x^1\) of the interval \([0,1]\). Let \(n>1\) iterations of the method be performed, during which trials were carried out at \(k = k(n)\) points \(x^i, 1\leqslant i\leqslant k\). Then the trial points  \(x^{k+1},\dotsc,x^{k+p}\) of the next \((n+1)\)-th iteration are determined in accordance with the following rules.

@@

Step 4. The characteristics \(R(i),1\leqslant i\leqslant k+1\), arrange in descending order 
\begin{displaymath}
\label{step4}
R(t_1)\geqslant R(t_2)\geqslant \dots \geqslant R(t_{k})\geqslant R(t_{k+1})
\end{displaymath}
and select \(p\) of the largest characteristics with interval numbers \(t_j,1\leqslant j\leqslant p\).

Step 5. Carry out new trials at the points \(x_{k+j},1\leqslant j\leqslant p\), calculated using the formulas
\begin{displaymath}
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2},\; t_j=1, \; t_j=k+1
\end{displaymath}
\begin{displaymath}
\label{step5}
{{x}^{k+1}}=\frac{{{x}_{{{t}_{j}}}}+{{x}_{{{t}_{j}}-1}}}{2}-\text{sign}({{z}_{{{t}_{j}}}}-{{z}_{{{t}_{j}}-1}})\frac{1}{2r}{{\left[ \frac{|{{z}_{{{t}_{j}}}}-{{z}_{{{t}_{j}}-1}}|}{\mu } \right]}^{N}}, 1<{{t}_{j}}<k+1.
\end{displaymath}

@@

Reducing multidimensional problems to one-dimensional ones through the use of evolvents has important properties such as continuity and preserving uniform bounding of function differences with limited argument variation. However, some information about the proximity of points in multidimensional space is lost, since the point \(x\in [0,1]\) has only left and right neighbors, and the corresponding point \(y(x) \in R^N\) has neighbors in \(2^N\) directions. As a result, when using evolvents, the images \(y' , y''\) that are close in \(N\)-dimensional space can correspond to rather distant preimages \(x' , x''\) on the interval \([0,1]\). This property leads to redundant calculations, because several limit points \(x' , x''\) of the test sequence generated by the method on the segment \([0,1]\), can correspond to a single limit point \(y\) in \(N\)-dimensional space.

A possible way to overcome this disadvantage is to use a set of evolvents (multiple mappings)
\begin{displaymath}
\label{eq:142}
Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{displaymath}
instead of using a single Peano curve \(y(x)\) (see \cite{Strongin1992,Strongin2000}). For example, each Peano curve \(y^i(x)\) from \(Y_L(x)\) can be obtained as a result of some shifting \(y(x)\) along the main diagonal of the hyperinterval \(D\). Another way is to rotate the evolvent \(y(x)\) around the origin. The set of evolvents that have been constructed allows us to obtain for any close images \(y', y''\) close preimages \(x', x''\) for some mapping \(y^i(x)\).

@@

However, each calculation of the value of the one-dimensional function  \(\varphi_1(y_1)\) at some fixed point corresponds to the solution of a one-dimensional minimization problem 
\begin{displaymath}
\label{phi2}
\varphi_2(y_1,y_2^*)=\min\{\varphi(y_1,y_2):y_2\in [a_2,b_2]\}.
\end{displaymath}
And so on, until the calculation   according to (\ref{phiN}).

@@

Using new variables, the main relation of the multistep scheme (\ref{nestedScheme}) can be rewritten as
\begin{equation}
\label{blockNested}
\min_{y\in D} \varphi(y)=\min_{u_1\in D_1}\min_{u_2\in D_2}\dots\min_{u_M\in D_M}\varphi(y),
\end{equation}
where the subdomains  \(D_i,1\leqslant i\leqslant M\), are projections of the original search domain \(D\) onto the subspaces corresponding to the variables \(u_1,1\leqslant i\leqslant M\).

The formulas that determine the method for solving problem (\ref{task}) based on relations (\ref{blockNested}) generally coincide with the recursive scheme (\ref{phiN})--(\ref{phiFirst}). One need only replace the original variables  \(y_i,1\leqslant i\leqslant N\), with block variables  \(u_1,1\leqslant i\leqslant M\). 
In this case, the fundamental difference from the original scheme is the fact that the block scheme has nested subproblems
\begin{equation}
\label{subTasks}
\varphi_i(u_1,\dots,u_i)=\min_{u_{i+1}\in D_{i+1}}\varphi_{i+1}(u_1,\dots,u_i,u_{i+1}),1\leqslant i\leqslant M-1,
\end{equation}
which are multidimensional, and to solve them, a method of reducing dimensionality based on Peano curves can be applied.

@@

The solution to the resulting set of subproblems (\ref{subTasks}) can be organized in various ways.
The obvious method (elaborated in detail in \cite{Grishagin2015} for the nested optimization scheme and in \cite{Barkalov2014} for the block nested optimization scheme) is based on solving subproblems in accordance with the recursive order of their generation. However, there is a significant loss of information about the target function.

@@

The adaptive dimensionality reduction scheme changes the order in which subproblems are solved: they will be solved not one by one (in accordance with their hierarchy in the task tree), but simultaneously, i.e., there will be a number of subproblems that are in the process of being solved. Under the new scheme:	
	\begin{itemize}
		\item to calculate the value of the function of the \(i\)-th level from (\ref{subTasks}), a new \(i+1\)-th level problem is generated in which trials are carried out, after which a new generated problem is included in the set of existing problems to be solved;
		\item the iteration of the global search consists in choosing \(p\) (the most promising) problem from the set of existing problems in which trials are carried out; points for new trials are determined in accordance with the parallel global search algorithm from Section \ref{Sec_GSA};
		\item the minimum values of functions from (\ref{subTasks}) are their current estimates based on the accumulated search information.
	\end{itemize}

@@

Therefore, at each iteration, subproblems with the maximum characteristic are selected for conducting the next trial. The trial either leads to the calculation of the value of the objective function \(\varphi_1(y)\) (if the selected subproblem belonged to the level \(j=M\)), or generates new subproblems according to (\ref{subTasks}) for \(j \geqslant  M-1\). In the latter case, the newly generated problems are added to the current set of problems, their characteristics are calculated, and the process is repeated. The optimization process is completed when the root problem satisfies the condition for stopping the algorithm that solves this problem. Some results pointing in this direction are presented in \cite{Barkalov2020}.

@@

In relation to global optimization methods, an operation that can be efficiently implemented on GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on GPU. Data transfers from CPU to GPU will be minimal: one need only transfer the coordinates of the trial points to GPU, and get back the function values at these points. Functions that determine the processing of trial results in accordance with the algorithm, and require work with large amounts of accumulated search information, can be efficiently implemented on CPU.

@@

To organize parallel calculations, we will use a set of evolvents and a block adaptive dimensionality reduction scheme. We take a small number of nesting levels, 2 to 3, in which the original large dimensionality problem is divided into 2-3 nested lower dimensional subproblems. We will use multiple evolvents only at the upper level of nesting, corresponding to the u1 variable block. This subproblem will be reduced to a set of one-dimensional problems that will be solved in parallel, each in a separate process. Trial results at point $x$ obtained for the problem solved by a specific processor are interpreted as the trial results in the remaining problems (at the corresponding points \(u^1_1,..., u^s_1\) ).
Then, applying an adaptive scheme to solve the nested subproblems (\ref{subTasks}), we get a parallel algorithm with a wide degree of variability.

Fig. \ref{fig:Parallel_Computing} shows the general scheme of organizing computations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using CPU only. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

The subproblem of the last \((M - 1)\)-th level 
\[
{{\varphi }_{M-1}}({{u}_{1}},...,{{u}_{M-1}})=\underset{{{u}_{M}}\in {{D}_{M}}}{\mathop{\min }}\,{{\varphi }_{M}}({{u}_{1}},...,{{u}_{M}})
\]
 differs from all the previous subproblems; it calculates the values of the objective function, since ${{\varphi }_{M}}({{u}_{1}},...,{{u}_{M}})=\varphi ({{y}_{1}},...,{{y}_{N}})$. This subproblem transfers between CPU and GPU.

@@

The numerical experiments were carried out using the Lomonosov supercomputer (Lomonosov Moscow State University). Each supercomputer node included two quad-core processors Intel Xeon X5570, two NVIDIA Tesla X2070 and 12 Gb RAM. When building the Globalizer system for running on the Lomonosov supercomputer, the GCC 4.3.0 compiler, CUDA 6.5 and Intel MPI 2017 were used.

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time of calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series). 
In the experiments carried out the average time for calculating the function values was 0.01 seconds, which exceeds the latency of the network or the data transfer time to the GPU.

@@

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA) from Section 2.  A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Gaviano2003}. The global minimum $y^*$ was considered found if the algorithm generated a trial point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\| \leqslant \delta $.  The size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved , $a$ and $b$ are the boundaries  of the search domain $D$, the parameter  $\Delta ={{10}^{-6}}$ for $N=4$ and $\Delta ={{10}^{-7}}$ for $N=5$. When using the GSA method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class $r=5.6$; the parameter for constructing the Peano curve was $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.

>>>>

\section{Multidimensional parallel global search algorithm}\label{Sec_GSA}
Let's consider the problem of finding the global minimum of the \(N\)-dimensional function \(\varphi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\).  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\varphi(y^*)=\min\{\varphi(y):y\in D\},
\end{equation}
\begin{equation}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty.
\end{equation}

@@

At the first iteration of the method, the trial is carried out at an arbitrary internal point \(x^1\) of the interval \([0,1]\). Let \(n>1\) iterations of the method be performed, during which trials were carried out at \(k = k(n)\) points \(x^i, 1\leqslant i\leqslant k\). Then the trial points  \(x^{k+1},\dotsc,x^{k+p}\) of the next \((n+1)\)th iteration are determined in accordance with the following rules.

@@

Step 4. Arrange characteristics \(R(i),1\leqslant i\leqslant k+1\), in descending order 
\begin{displaymath}
\label{step4}
R(t_1)\geqslant R(t_2)\geqslant \dots \geqslant R(t_{k})\geqslant R(t_{k+1})
\end{displaymath}
and select \(p\) of the largest characteristics with interval numbers \(t_j,1\leqslant j\leqslant p\).

Step 5. Carry out new trials at the points \(x_{k+j},1\leqslant j\leqslant p\), calculated using the formulas
\begin{displaymath}
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2},\; t_j=1, \; t_j=k+1,
\end{displaymath}
\begin{displaymath}
\label{step5}
{{x}^{k+1}}=\frac{{{x}_{{{t}_{j}}}}+{{x}_{{{t}_{j}}-1}}}{2}-\text{sign}({{z}_{{{t}_{j}}}}-{{z}_{{{t}_{j}}-1}})\frac{1}{2r}{{\left[ \frac{|{{z}_{{{t}_{j}}}}-{{z}_{{{t}_{j}}-1}}|}{\mu } \right]}^{N}}, 1<{{t}_{j}}<k+1.
\end{displaymath}

@@

Reducing multidimensional problems to one-dimensional ones through the use of evolvents has important properties such as continuity and preserving uniform bounding of function differences with limited argument variation. However, some information about the proximity of points in multidimensional space is lost, since the point \(x\in [0,1]\) has only left and right neighbors, and the corresponding point \(y(x) \in R^N\) has neighbors in \(2^N\) directions. As a result, when using evolvents, the images \(y' , y''\) that are close in \(N\)-dimensional space can correspond to rather distant preimages \(x' , x''\) on the interval \([0,1]\). This property leads to redundant calculations, because several limit points \(x' , x''\) of the sequence of the trial points generated by the method on the segment \([0,1]\), can correspond to a single limit point \(y\) in \(N\)-dimensional space.

A possible way to overcome this disadvantage is to use a set of evolvents (\textit{multiple mappings})
\begin{displaymath}
\label{eq:142}
Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{displaymath}
instead of using a single Peano curve \(y(x)\) (see \cite{Strongin1992,Strongin2000}). For example, each Peano curve \(y^i(x)\) from \(Y_L(x)\) can be obtained as a result of some shifting \(y(x)\) along the main diagonal of the hyperinterval \(D\). Another way is to rotate the evolvent \(y(x)\) around the origin. The set of evolvents that have been constructed allows us to obtain for any close images \(y', y''\) close preimages \(x', x''\) for some mapping \(y^i(x)\).

@@

However, each calculation of the value of the one-dimensional function  \(\varphi_1(y_1)\) at some fixed point corresponds to the solution of a one-dimensional minimization problem 
\begin{displaymath}
\label{phi2}
\varphi_2(y_1,y_2^*)=\min\{\varphi(y_1,y_2):y_2\in [a_2,b_2]\}.
\end{displaymath}
And so on, until the calculation of $\varphi_N$  according to (\ref{phiN}).

@@

Using new variables, the main relation of the nested scheme (\ref{nestedScheme}) can be rewritten as
\begin{equation}
\label{blockNested}
\min_{y\in D} \varphi(y)=\min_{u_1\in D_1}\min_{u_2\in D_2}\dots\min_{u_M\in D_M}\varphi(y),
\end{equation}
where the subdomains  \(D_i,1\leqslant i\leqslant M\), are projections of the original search domain \(D\) onto the subspaces corresponding to the variables \(u_1,1\leqslant i\leqslant M\).

The formulas that determine the method for solving problem (\ref{task}) based on relations (\ref{blockNested}) generally coincide with the recursive scheme (\ref{phiN})--(\ref{phiFirst}). One need only replace the original variables  \(y_i,1\leqslant i\leqslant N\), with block variables  \(u_1,1\leqslant i\leqslant M\). 
In this case, the fundamental difference from the original scheme is the fact that the block scheme has nested subproblems
\begin{equation}
\label{subTasks}
\varphi_i(u_1,\dots,u_i)=\min_{u_{i+1}\in D_{i+1}}\varphi_{i+1}(u_1,\dots,u_i,u_{i+1}),1\leqslant i\leqslant M-1,
\end{equation}
which are multidimensional, and to solve them a method of reducing dimensionality based on Peano curves can be applied.

@@

Solving the resulting set of subproblems (\ref{subTasks}) can be organized in various ways.
The obvious method (elaborated in detail in \cite{Grishagin2015} for the nested optimization scheme and in \cite{Barkalov2014} for the block nested optimization scheme) is based on solving subproblems in accordance with the recursive order of their generation. However, in this case there is a significant loss of information about the target function.

@@

The adaptive dimensionality reduction scheme changes the order in which subproblems are solved: they will be solved not one by one (in accordance with their hierarchy in the task tree), but simultaneously, i.e., there will be a number of subproblems that are in the process of being solved. Under the new scheme:	
	\begin{itemize}
		\item to calculate the value of the function of the \(i\)-th level from (\ref{subTasks}), a new \((i+1)\)th level problem is generated in which trials are carried out, after which a new generated problem is included in the set of existing problems to be solved;
		\item the iteration of the global search consists in choosing \(p\) (the most promising) problems from the set of existing problems in which trials are carried out; points for new trials are determined in accordance with the parallel global search algorithm from Section \ref{Sec_GSA};
		\item the minimum values of functions from (\ref{subTasks}) are their current estimates based on the accumulated search information.
	\end{itemize}

@@

Therefore, at each iteration, subproblems with the maximum characteristic are selected for conducting the next trial. The trial either leads to the calculation of the value of the objective function \(\varphi_1(y)\) (if the selected subproblem belonged to the level \(j=M\)), or generates new subproblems according to (\ref{subTasks}) for \(j \leqslant  M-1\). In the latter case, the newly generated problems are added to the current set of problems, their characteristics are calculated, and the process is repeated. The optimization process is completed when the root problem satisfies the condition for stopping the algorithm that solves this problem. Some results pointing in this direction are presented in \cite{Barkalov2020}.

@@

In relation to global optimization methods, an operation that can be efficiently implemented on GPU is the parallel calculation of many values of the objective function at once. Naturally, this requires implementing a procedure for calculating the value of a function on GPU. Data transfers from CPU to GPU will be minimal: one need only transfer the coordinates of the trial points to GPU, and get back the function values at these points. Functions that determine the processing of trial results in accordance with the algorithm, and require work with large amount of accumulated search information, can be efficiently implemented on CPU.

@@

To organize parallel calculations, we will use a set of evolvents and a block adaptive dimensionality reduction scheme. We take a small number of nesting levels, in which the original large dimensionality problem is divided into 2-3 nested lower dimensional subproblems. We will use multiple evolvents only at the upper level of nesting, corresponding to the variable $u_1$. This subproblem will be reduced to a set of one-dimensional problems that will be solved in parallel, each in a separate process. Trial results at point $x$ obtained for the problem solved by a specific processor are interpreted as the trial results in the remaining problems (at the corresponding points \(u^1_1,..., u^s_1\) ).
Then, applying an adaptive scheme to solve the nested subproblems (\ref{subTasks}), we get a parallel algorithm with a wide degree of variability.

Fig. \ref{fig:Parallel_Computing} shows the general scheme of organizing computations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using CPU only. The values of the function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.
The subproblem of the last \((M-1)\)-th level 
\[
{{\varphi }_{M-1}}({{u}_{1}},...,{{u}_{M-1}})=\underset{{{u}_{M}}\in {{D}_{M}}}{\mathop{\min }}\,{{\varphi }_{M}}({{u}_{1}},...,{{u}_{M}})
\]
 differs from all the previous subproblems; it calculates the values of the objective function, since ${{\varphi }_{M}}({{u}_{1}},...,{{u}_{M}})=\varphi ({{y}_{1}},...,{{y}_{N}})$. This subproblem transfers data between CPU and GPU.

@@

The numerical experiments were carried out using the Lomonosov supercomputer (Lomonosov Moscow State University). Each supercomputer node included two quad-core processors Intel Xeon X5570, two NVIDIA Tesla X2070 and 12 Gb RAM. To build the program for running on the Lomonosov supercomputer, the GCC 4.3.0 compiler, CUDA 6.5 and Intel MPI 2017 were used.

Note that well-known test problems from the field of multidimensional global optimization are characterized by a short time of calculating the values of the objective function. Therefore, in order to simulate the computational complexity inherent in applied optimization problems \cite{Modorskii2016}, the calculation of the objective function in all experiments was complicated by additional calculations that do not change the form of the function and the location of its minima (summing the segment of the Taylor series). 
In the experiments carried out the average time for calculating the function values was 0.01 seconds, which exceeds the latency of the network or the data transfer time between CPU and GPU.

@@

Below are the results of a numerical comparison of three sequential algorithms: DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA) from Section 2.  A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Gaviano2003}. The global minimum $y^*$ was considered found if the algorithm generated a trial point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\| \leqslant \delta $.  The size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved, $a$ and $b$ are the boundaries  of the search domain $D$, the parameter  $\Delta ={{10}^{-6}}$ for $N=4$ and $\Delta ={{10}^{-7}}$ for $N=5$. When using the GSA method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class $r=5.6$; the parameter for constructing the Peano curve was $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.
